Supplementary Material for `The response of trophic interaction networks to multiple stressors in a marine latitudinal gradient of the Southern Hemisphere'

### 2024-06-10 manuscript.Rmd
Updated manuscript with minor revisions

Ecological networks offer valuable insights into community structure, key species identification, and ecosystem management. Understanding how these networks respond to global change stressors is of increasing interest, especially along geographical gradients.
This review summarizes potential stressor responses in marine food webs from the Southwest Atlantic to the Antarctic (45 - 78ºS), encompassing areas such as San Jorge Gulf, Beagle Channel, Burdwood Bank, Scotia Sea, Potter Cove, and the Weddell Sea in Antarctica. The objectives are: 1) to describe the structure of marine food webs along this latitudinal axis using a network approach; 2) to identify predominant global change-related stressors affecting each ecosystem; and 3) to summarize observed food web changes and hypothesize on stressor impacts. The effects of stressors were primarily reviewed at the species level.
Alternative hypotheses for each study area were formulated considering: a) main stressors; b) impacted parameters; c) node-level species properties; and d) network-level food web properties. Global warming emerges as the most common stressor among the studied areas across the latitudinal gradient, except in the Beagle Channel and Burdwood Bank, where alien species introduction and fisheries are more influential. We offer a series of alternative hypotheses on how warming may affect the food webs. This review emphasizes the benefits of using a network approach to understand and predict stressor effects in Southern Hemisphere marine ecosystems. This approach provides a holistic understanding of ecosystems, which enhances our ability to identify key species and their interactions, offering insights for ecosystem management and conservation in the face of global change stressors.

@@

Here we review for the first time stressor responses of marine food webs for areas located along a Southwest Atlantic to Antarctic gradient (45 - 78ºS, Figure 1). We considered proven and expected changes in food webs driven by global change-related stressors in selected areas along this large-scale latitudinal range.
The aim of this review is threefold: 1) describe the complexity and structure of marine food webs along the Southwest Atlantic to Antarctic axis from a network perspective; 2) identify the ongoing global change-related stressors for each marine ecosystem containing the food webs; and 3) review proven food web changes and elaborate on a series of alternative hypotheses on how the identified stressors might affect food web features (energy flow, stability and topology), combining information on node- and network-level properties. 
To identify the main stressors, we performed an exploratory and configurative review instead of a systematic review [@Haddaway2015], because our objective was to formulate evidenced-based hypotheses rather than providing an exhaustive list of species-level stressors for each ecosystem.
Finally, we suggest which additional data and analyses are needed to gain insights into the stressors’ effects on food web properties in the Southwest Atlantic to Antarctic region.

________________

@@

The Southwest and the Atlantic sectors of the Southern Ocean comprise one of the most biologically productive regions of the world's oceans [@Acha2004; @Latorre2023]. The referred region extends from San Jorge Gulf (45ºS) in the Patagonian shelf to the Weddell Sea (78ºS) in the Southern Ocean and covers a well-connected oceanic area [@Matano2010; @Guihou2020].

Throughout this region, many studies address the ecology of predator-prey relationships but most of these studies are restricted to a few interacting species or certain components of the food web such as fish, crustaceans, and benthos (e.g., @Vinuesa2007, @Padovani2012, @Pasotti2015, @Riccialdelli2020, @Saunders2019, @Pasti2021, @Cordone2022). Only few studies consider the complexity of trophic interactions within highly resolved food webs containing hundreds of species, ranging from phytoplankton, zooplankton, benthos, fish, seabirds to whales (but see @Jacob2011, @Marina2018, @Funes2022, @Lopez-Lopez2022, @Rodriguez2022, @Marinainrev). Neglecting this complexity might lead to a misunderstanding of the structure and functioning of the ecosystems and ultimately reduce the ability to predict ecosystem responses to perturbations [@Montoya2009].

@@

In this review, we included six large marine areas in the Southern Hemisphere for which highly-resolved empirical food webs in terms of species diversity and trophic interactions have been previously described: (1) San Jorge Gulf (~45 - 47ºS, 65 - 68ºW) [@Funes2022], (2) Beagle Channel (~54ºS, 68ºW) [@Rodriguez2022], (3) Burdwood Bank (~54ºS, 59ºW) [@Marinainrev], (4) Scotia Sea (~48 - 58ºS, 50ºW) [@Lopez-Lopez2022], (5) Potter Cove (62ºS, 58ºW, Antarctica) [@Marina2018], and (6) Weddell Sea (74 - 78ºS, 30ºW, Antarctica) [@Jacob2011; @Marina2024] (Figure 1). The geographic locations of these marine biomes span from temperate Atlantic waters to Antarctic, and are exposed to global change stressors (e.g., sea warming, glacial retreat, fishery, pollution) [@Sage2020], of which some are locally more important, e.g., fisheries in the San Jorge Gulf or glacial retreat in the Antarctic.

@@

Located at a more southern latitude (~54ºS, 68ºW), the Beagle Channel (Figure 1) is an interoceanic passage at the southernmost tip of South America, spanning 240 km in length and 5 km in width (~1200 $\text{km}^2$), with a depth range of 0 - 140 m. It features complex coastlines, varying bathymetry, a prevailing west-to-east circulation pattern, and a significant longitudinal gradient of glacial freshwater discharge [@Schloss2023]. @Bruno2023 suggests that locally produced suspended particulate organic matter (mostly composed of phytoplankton) and organic matter accumulated in the sediments (macroalgae-originated detritus) are the primary food sources for the marine ecosystem, as opposed to allochthonous materials.

@@

The so-called Burdwood Bank ecosystem comprises Marine Protected Areas Namuncurá - Burdwood Bank I and II (~54ºS, 59ºW), meaning the shallow submarine plateau named Burdwood Bank with a 200 m isobath boundary, and a deep slope that reaches 4000 m in depth, respectively [@APN2022] (Figure 1). Physical features in the plateau are fairly stable, with salinity averaging 34 all year round and temperature ranging between 4 and 8ºC [@Acha2004]. The plateau is surrounded by steep flanks of up to 4000 m depth, protected by the Marine Protected Area Namuncurá - Burdwood Bank II (32,000 $\text{km}^2$). Intense upwelling and mixing occur in relation to the slope, entraining deep nutrient-rich waters into the photic layer [@Matano2019], and resulting in a fairly homogeneous water column both spatially and temporally [@Matano2019].

The Burdwood Bank food web comprises 379 nodes and 1788 interactions, with a connectance of 0.01, and an asymmetric degree distribution (i.e. most of the species have a relatively low number of interactions and few species concentrate most of them). Almost half of the consumers are omnivores (48%), and the network displays a small-world pattern [@Marinainrev] (Table 1). This network pattern implies a rapid spread of a perturbation (e.g., contaminant, population fluctuation, local extinction) throughout the food web due to a short distance among species but, at the same time, a greater resistance caused by the confinement of perturbations mainly within subnetworks as a result of a high clustering coefficient [@Bornatowski2017; @Dormann2017].

@@

The Scotia Sea is a deep-sea basin (~48 - 58ºS, 50ºW), delimited by the Drake Passage to the West and by the island complex of the Scotia Arc to the North, East, and South, with an approximate extension of 1.5 x 106 $\text{km}^2$ and a depth range of 0 - 3000 m [@Murphy2006] (Figure 1). Its oceanography is dominated by the Antarctic Circumpolar Current, which is spatially structured by frontal systems [@Whitworth1980]. The South Antarctic Circumpolar Current Front subdivides the Scotia Sea into two biogeographic regions: the Northern Scotia Sea, characterized by higher and more variable temperatures, and the Southern Scotia Sea, characterized by lower and more stable temperatures and influenced by seasonal sea ice [@Raymond2011].

@@

Located between 74 and 78ºS, the high Antarctic Weddell Sea shelf spans approximately 450 km from East to West [@Jacob2011] (Figure 1). The water depth ranges from 200 to 500 m, with shallower regions being covered by continental ice that forms the coastline along the eastern and southern parts of the Weddell Sea. The shelf area is characterized by a complex three-dimensional benthic habitat, substantial benthic biomasses, and an intermediate to high diversity [@Teixido2002].

>>>>

Ecological networks offer valuable insights into community structure, key species identification, and ecosystem management. Understanding how these networks respond to global change stressors is of increasing interest, especially along geographical gradients.
This review summarizes potential stressor responses in marine food webs from the Southwest Atlantic to the Antarctic (45 - 78°S), encompassing areas such as San Jorge Gulf, Beagle Channel, Burdwood Bank, Scotia Sea, Potter Cove, and the Weddell Sea in Antarctica. The objectives are: 1) to describe the structure of marine food webs along this latitudinal axis using a network approach; 2) to identify predominant global change-related stressors affecting each ecosystem; and 3) to summarize observed food web changes and hypothesize on stressor impacts. The effects of stressors were primarily reviewed at the species level.
Alternative hypotheses for each study area were formulated considering: a) main stressors; b) impacted parameters; c) node-level species properties; and d) network-level food web properties. Global warming emerges as the most common stressor among the studied areas across the latitudinal gradient, except in the Beagle Channel and Burdwood Bank, where alien species introduction and fisheries are more influential. We offer a series of alternative hypotheses on how warming may affect the food webs. This review emphasizes the benefits of using a network approach to understand and predict stressor effects in Southern Hemisphere marine ecosystems. This approach provides a holistic understanding of ecosystems, which enhances our ability to identify key species and their interactions, offering insights for ecosystem management and conservation in the face of global change stressors.

@@

Here we review for the first time stressor responses of marine food webs for areas located along a Southwest Atlantic to Antarctic gradient (45 - 78°S, Figure 1). We considered proven and expected changes in food webs driven by global change-related stressors in selected areas along this large-scale latitudinal range.
The aim of this review is threefold: 1) describe the complexity and structure of marine food webs along the Southwest Atlantic to Antarctic axis from a network perspective; 2) identify the ongoing global change-related stressors for each marine ecosystem containing the food webs; and 3) review proven food web changes and elaborate on a series of alternative hypotheses on how the identified stressors might affect food web features (energy flow, stability and topology), combining information on node- and network-level properties. 
To identify the main stressors, we performed an exploratory and configurative review instead of a systematic review [@Haddaway2015], where we considered more than 200 peer-reviewed articles retrieved mainly from Google Scholar using keywords combining location ( "San Jorge Gulf", "Beagle Channel", "Burdwood Bank", "Scotia Sea", "Weddell Sea", "Potter Cove") and stressors ("acidification", "alien species", "climate change", "contamination", "fishery", "glacier retreat", "global warming", "ice scouring", "invasive species", "marine contaminant", "microplastics", "mercury", "pollution", "sea ice", "sea warming", "urban waste"). Additionally, we specifically searched for publications by authors we know may have conducted related studies in the area. We performed this type of review because our objective was to formulate evidenced-based hypotheses rather than providing an exhaustive list of species-level stressors for each ecosystem.
Finally, we suggest which additional data and analyses are needed to gain insights into the stressors’ effects on food web properties in the Southwest Atlantic to Antarctic region.

@@

The Southwest and the Atlantic sectors of the Southern Ocean comprise one of the most biologically productive regions of the world's oceans [@Acha2004; @Latorre2023]. The referred region extends from San Jorge Gulf (45°S) in the Patagonian shelf to the Weddell Sea (78°S) in the Southern Ocean and covers a well-connected oceanic area [@Matano2010; @Guihou2020].

Throughout this region, many studies address the ecology of predator-prey relationships but most of these studies are restricted to a few interacting species or certain components of the food web such as fish, crustaceans, and benthos (e.g., @Vinuesa2007, @Padovani2012, @Pasotti2015, @Riccialdelli2020, @Saunders2019, @Pasti2021, @Cordone2022). Only few studies consider the complexity of trophic interactions within highly resolved food webs containing hundreds of species, ranging from phytoplankton, zooplankton, benthos, fish, seabirds to whales (but see @Jacob2011, @Marina2018, @Funes2022, @Lopez-Lopez2022, @Rodriguez2022, @Marina2024a). Neglecting this complexity might lead to a misunderstanding of the structure and functioning of the ecosystems and ultimately reduce the ability to predict ecosystem responses to perturbations [@Montoya2009].

@@

In this review, we included six large marine areas in the Southern Hemisphere for which highly-resolved empirical food webs in terms of species diversity and trophic interactions have been previously described: (1) San Jorge Gulf (~45 - 47°S, 65 - 68°W) [@Funes2022], (2) Beagle Channel (~54°S, 68°W) [@Rodriguez2022], (3) Burdwood Bank (~54°S, 59°W) [@Marina2024a], (4) Scotia Sea (~48 - 58°S, 50°W) [@Lopez-Lopez2022], (5) Potter Cove (62°S, 58°W, Antarctica) [@Marina2018], and (6) Weddell Sea (74 - 78°S, 30°W, Antarctica) [@Jacob2011; @Marina2024] (Figure 1). The geographic locations of these marine biomes span from temperate Atlantic waters to Antarctic, and are exposed to global change stressors (e.g., sea warming, glacial retreat, fishery, pollution) [@Sage2020], of which some are locally more important, e.g., fisheries in the San Jorge Gulf or glacial retreat in the Antarctic.

@@

Located at a more southern latitude (~54°S, 68°W), the Beagle Channel (Figure 1) is an interoceanic passage at the southernmost tip of South America, spanning 240 km in length and 5 km in width (~1200 $\text{km}^2$), with a depth range of 0 - 140 m. It features complex coastlines, varying bathymetry, a prevailing west-to-east circulation pattern, and a significant longitudinal gradient of glacial freshwater discharge [@Schloss2023]. @Bruno2023 suggests that locally produced suspended particulate organic matter (mostly composed of phytoplankton) and organic matter accumulated in the sediments (macroalgae-originated detritus) are the primary food sources for the marine ecosystem, as opposed to allochthonous materials.

@@

The so-called Burdwood Bank ecosystem comprises Marine Protected Areas Namuncurá - Burdwood Bank I and II (~54°S, 59°W), meaning the shallow submarine plateau named Burdwood Bank with a 200 m isobath boundary, and a deep slope that reaches 4000 m in depth, respectively [@APN2022] (Figure 1). Physical features in the plateau are fairly stable, with salinity averaging 34 all year round and temperature ranging between 4 and 8°C [@Acha2004]. The plateau is surrounded by steep flanks of up to 4000 m depth, protected by the Marine Protected Area Namuncurá - Burdwood Bank II (32,000 $\text{km}^2$). Intense upwelling and mixing occur in relation to the slope, entraining deep nutrient-rich waters into the photic layer [@Matano2019], and resulting in a fairly homogeneous water column both spatially and temporally [@Matano2019].

The Burdwood Bank food web comprises 379 nodes and 1788 interactions, with a connectance of 0.01, and an asymmetric degree distribution (i.e. most of the species have a relatively low number of interactions and few species concentrate most of them). Almost half of the consumers are omnivores (48%), and the network displays a small-world pattern [@Marina2024a] (Table 1). This network pattern implies a rapid spread of a perturbation (e.g., contaminant, population fluctuation, local extinction) throughout the food web due to a short distance among species but, at the same time, a greater resistance caused by the confinement of perturbations mainly within subnetworks as a result of a high clustering coefficient [@Bornatowski2017; @Dormann2017].

@@

The Scotia Sea is a deep-sea basin (~48 - 58°S, 50°W), delimited by the Drake Passage to the West and by the island complex of the Scotia Arc to the North, East, and South, with an approximate extension of 1.5 x 106 $\text{km}^2$ and a depth range of 0 - 3000 m [@Murphy2006] (Figure 1). Its oceanography is dominated by the Antarctic Circumpolar Current, which is spatially structured by frontal systems [@Whitworth1980]. The South Antarctic Circumpolar Current Front subdivides the Scotia Sea into two biogeographic regions: the Northern Scotia Sea, characterized by higher and more variable temperatures, and the Southern Scotia Sea, characterized by lower and more stable temperatures and influenced by seasonal sea ice [@Raymond2011].

@@

Located between 74 and 78°S, the high Antarctic Weddell Sea shelf spans approximately 450 km from East to West [@Jacob2011] (Figure 1). The water depth ranges from 200 to 500 m, with shallower regions being covered by continental ice that forms the coastline along the eastern and southern parts of the Weddell Sea. The shelf area is characterized by a complex three-dimensional benthic habitat, substantial benthic biomasses, and an intermediate to high diversity [@Teixido2002].
